---
title: "ggplot Tutorial"
format: html
editor: visual
---

## ggplot2 - The 'Burger' of the R coding Universe

-   Always some 'bread' and a 'patty'

-   Can add salad and sauces to taste.

-   Build your burger in layers(!)

## Install those packages!

```{r}
library(tidyverse)
```

## Grab your data

R includes loads of amazing datasets already pre-loaded in. You can see all available datasets by running the code below.

```{r}
data()
```

Let's load in mtcars:

```{r}
data(mtcars)
```

## LUMOS!

View the first few rows of the dataset:

```{r}
head(mtcars)
```

The all powerful 'View()' function

```{r}
View(mtcars)
```

## Let's ggplot!

### The basic burger:

-   Get your bread

-   Choose your patty

-   Multiple pattys?

```{r}
ggplot(data = mtcars, aes(x = wt, y = mpg)) + # here's our bread!
  geom_point() # here's our chosen patty 
  
```

## Let's add some salad

### Titles and Labels

```{r}
ggplot(data = mtcars, aes(x = wt, y = mpg)) + # here's our bread!
  geom_smooth(method = "loess") + # here's our chosen patty 
  geom_point() + # here's an additional patty
  labs( # here's some salad items
  )
```

## Let's add some sauce

-   Color by Cylinder

```{r, warning = FALSE}
ggplot(data = mtcars, aes(x = wt, y = mpg)) + # here's our bread!
  geom_smooth(method = "loess") + # here's our chosen patty 
  geom_point() + # here's an additional patty
  labs( # here's some salad items
    title = "Scatter Plot of Weight vs MPG",
    x = "Weight",
    y = "Miles per Gallon (MPG)"
  )
```

## Add some seasoning

-   applying themes

```{r, warning = FALSE}
ggplot(data = mtcars, aes(x = wt, y = mpg, color = cyl)) + # here's our wheat loaf bread!
  geom_smooth(method = "loess") + # here's our chosen patty 
  geom_point() + # here's an additional patty
  labs( # here's some salad items
    title = "Scatter Plot of Weight vs MPG",
    x = "Weight",
    y = "Miles per Gallon (MPG)",
    color = "Cylinders"
  )
```

## Let's make multiple burgers!

-   Use 'faceting' chef

```{r}
ggplot(data = mtcars, aes(x = wt, y = mpg)) + # here's our bread
  geom_point() + # here's our patty
  geom_smooth() + # here's our second patty
  facet_wrap(~ cyl) # implement our facet chef
```

## Wizard chef!

-   Introducing plotly

```{r, messages = FALSE}
library(plotly)

plot <- ggplot(data = mtcars, aes(x = wt, y = mpg, color = cyl)) + # here's our wheat loaf bread!
  geom_smooth(method = "loess") + # here's our chosen patty 
  geom_point() + # here's an additional patty
  labs( # here's some salad items
    title = "Scatter Plot of Weight vs MPG",
    x = "Weight",
    y = "Miles per Gallon (MPG)",
    color = "Cylinders"
  ) +
  theme_bw() # here's our seasoning

ggplotly(plot)
```

## Serving up

-   ggsave

```{r}
plot <- ggplot(data = mtcars, aes(x = wt, y = mpg, color = cyl)) + # here's our wheat loaf bread!
  geom_smooth(method = "loess") + # here's our chosen patty 
  geom_point() + # here's an additional patty
  labs( # here's some salad items
    title = "Scatter Plot of Weight vs MPG",
    x = "Weight",
    y = "Miles per Gallon (MPG)",
    color = "Cylinders"
  ) +
  theme_bw() # here's our seasoning

ggsave(
  plot = plot, 
  filename = file.path('42_ggplot_tutorial/plot.png')
)
```
