---
title: "Correlation: A love story between two variables that shouldn't be dating"
format: html
---

# Warm-up: Guess the correlation

- Positive / Negative Correlation?
- Strong / Weak Correlation?
- Suggest a number?

## Example 1
```{r}
set.seed(1)
ex1 <- dplyr::tibble(
  x = rnorm(100),
  y = -2*x + rnorm(100, sd = 0.3)
)

ex1 |> 
  ggplot2::ggplot(ggplot2::aes(x, y)) +
  ggplot2::geom_point(size = 2) +
  ggplot2::labs(title = "Example 1") +
  ggplot2::theme(text = ggplot2::element_text(size = 16))
```

## Example 2
```{r}
set.seed(1)
ex2 <- dplyr::tibble(
  x = seq(0, 2, length = 100),
  y = x^10 + rnorm(100, sd = 0.01)
)

ex2 |> 
  ggplot2::ggplot(ggplot2::aes(x, y)) +
  ggplot2::geom_point(size = 2) +
  ggplot2::labs(title = "Example 2") +
  ggplot2::theme(text = ggplot2::element_text(size = 16))
```

## Example 3
```{r}
set.seed(1)
ex3 <- dplyr::tibble(
  x = rnorm(100),
  y = -x + rnorm(100, sd = 1)
) |> 
  dplyr::add_row(x = 10, y = 10)

ex3 |> 
  ggplot2::ggplot(ggplot2::aes(x, y)) +
  ggplot2::geom_point(size = 2) +
  ggplot2::labs(title = "Example 3") +
  ggplot2::theme(text = ggplot2::element_text(size = 16))

```

## Example 4
```{r}
set.seed(1)
ex4 <- dplyr::tibble(
  x = rnorm(100),
  y = 0.01*x + rnorm(100, sd = 0.01)
)

ex4 |> 
  ggplot2::ggplot(ggplot2::aes(x, y)) +
  ggplot2::geom_point(size = 2) +
  ggplot2::labs(title = "Example 4") +
  ggplot2::lims(y = c(-0.5, 0.5)) +
  ggplot2::theme(text = ggplot2::element_text(size = 16))

```

# Introducing Spearman and Pearson

## Example 1
```{r}
ex1 |> 
    dplyr::summarise(
        example = 1,
        pearson = cor(x, y, method = "pearson"),
        spearman = cor(x, y, method = "spearman")
    )
```

## Example 2
```{r}
ex2 |> 
    dplyr::summarise(
        example = 2,
        pearson = cor(x, y, method = "pearson"),
        spearman = cor(x, y, method = "spearman")
    )
```

## Example 3
```{r}
ex3 |> 
    dplyr::summarise(
        example = 3,
        pearson = cor(x, y, method = "pearson"),
        spearman = cor(x, y, method = "spearman")
    )
```

## Example 4
```{r}
ex4 |> 
    dplyr::summarise(
        example = 4,
        pearson = cor(x, y, method = "pearson"),
        spearman = cor(x, y, method = "spearman")
    )
```

## Key Takeaways

Pearson's Correlation cares about linearity.
Spearman's Correlation cares about order.

## So when should I use each?

Use Pearson's if:
- Your data is linear (straight line trend)
- Your data is normally distributed
- Your variables are continuous

Use Spearman's if:
- Your data is not linear
- Your relationship is monotonic (always increasing or decreasing)
- Your data contains outliers that may distore Pearsons
- Your data is ordinal (e.g. ranked data)

# Real Data Example

```{r}
mtcars_tbl <- mtcars |> 
  dplyr::as_tibble(rownames = "model")
```

```{r}
mtcars_tbl |> 
  ggplot2::ggplot(ggplot2::aes(hp, mpg)) +
  ggplot2::geom_point(size = 2) +
  #ggplot2::geom_smooth(method = "lm", se = FALSE, colour = "red") +
  ggplot2::labs(title = "Horsepower vs MPG") +
  ggplot2::theme(text = ggplot2::element_text(size = 16))
```


```{r}
mtcars_tbl |> 
  dplyr::summarise(
    pearson  = cor(hp, mpg, method = "pearson"),
    spearman = cor(hp, mpg, method = "spearman")
  )
```

## Full correlation matrix

```{r}
numeric_vars <- mtcars_tbl |> 
    dplyr::select(dplyr::where(is.numeric))

cor_mat <- numeric_vars |> 
    cor(method = "pearson") |> 
    as.data.frame() |> 
    tibble::rownames_to_column("var1") |> 
    tidyr::pivot_longer(-var1, names_to = "var2", values_to = "correlation")
```

```{r}
cor_mat |> 
  ggplot2::ggplot(ggplot2::aes(var1, var2, fill = correlation)) +
  ggplot2::geom_tile() +
  ggplot2::scale_fill_gradient2(limits = c(-1, 1)) +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1)) +
  ggplot2::labs(title = "Correlation matrix")
```


## Extra Challenges (if there's time)
Pick two numerical variables from mtcars_tbl and decide whether to use Spearman's or Pearson's

(Not MPG vs HP)

```{r}
mtcars_tbl |> 
  ggplot2::ggplot(ggplot2::aes(cyl, disp)) +
  ggplot2::geom_point(size = 2) +
  ggplot2::geom_smooth(method = "lm", se = FALSE, colour = "red") +
  ggplot2::theme(text = ggplot2::element_text(size = 16))

mtcars_tbl |> 
  dplyr::summarise(
    pearson  = cor(cyl, disp, method = "pearson"),
    spearman = cor(cyl, disp, method = "spearman")
  )
```

# Bonus: Correlation != Causation
https://www.tylervigen.com/spurious-correlations