---
title: "Let's Talk About Data"
format: html
---

```{r}
# Create a small dataset
data <- data.frame(
    Name = c( # Text
        "Alice","Bob","Charlie","Diana","Ethan",
        "Fiona","George","Hannah","Ian","Julia"
    ), 
    Gender = c( # Nominal Categorical
        "Female","Male","Male","Female","Male",
        "Female","Male","Female","Male","Female"
    ), 
    Education = c( # Ordinal Categorical
        "Bachelor","Master","PhD","High School","Bachelor",
        "Master","PhD","High School","Bachelor","Master"
    ), 
    Department = c( # Nominal Categorical
        "HR","Finance","Engineering","Marketing","Engineering",
        "Finance","HR","Marketing","Engineering","Finance"
    ),  
    Age = c(23, 35, 29, 42, 31, 27, 38, 19, 45, 33),  # Continuous / Discrete Numerical
    Children = c(0,2,1,3,2,0,1,0,4,2),  # Discrete Numerical
    PassedExam = c( # Binary
        TRUE, TRUE, FALSE, TRUE, FALSE,
        TRUE, TRUE, FALSE, TRUE, TRUE
    ),  
    Height_cm = c( # Continuous Numerical
        165.2, 178.5, 172.0, 160.3, 182.1,
        168.4, 175.0, 158.9, 180.2, 170.5
    )  
)

# Preview the dataset
View(data)
```

## Types of Data

In every dataset we work with, the type of data determines everything that follows:

- Which statistical methods we use
- How we visualise it

Broadly speaking, we can split data into two big families:

1. Numerical - Data represents quantities (e.g. age)
2. Categorical - Data represents groups or labels (e.g. gender)

Why does it matter?
If we misclassify our data, we risk using the wrong statistical test and/or drawing misleading conclusions!

### Numerical

- Height of everyone who works at City and Guilds
- Average daily temperature in the UK
- Number of ice cream sales 
- Marks acheived by candidates in exams

Numerical Data can be split into two sub types: 
1. Discrete - e.g. count data, data cannot be a decimal
2. Continuous - e.g. measure data, data can be a decimal

#### Discrete

```{r}
# Frequency table
data |> 
    dplyr::count(Children)

# Calculate the median number of children
data |> 
    dplyr::summarise(
        median = median(Children)
    )

# Bar plot
ggplot2::ggplot(
    data = data,
    ggplot2::aes(x = Children)
) +
ggplot2::geom_bar() +
ggplot2::labs(
    title = "Number of Children",
    x = "Children",
    y = "Frequency"
)
```

#### Continuous

```{r}
# 
data |> 
    dplyr::summarise(
        mean_height = mean(Height_cm),
        sd_height = sd(Height_cm),
        min_height = min(Height_cm),
        max_height = max(Height_cm)
    )

# Histogram
ggplot2::ggplot(
    data = data,
    ggplot2::aes(x = Height_cm)
) +
ggplot2::geom_histogram(
    bins = 5,
    colour = "black"
) +
ggplot2::labs(
    title = "Distribution of Heights",
    x = "Height (cm)",
    y = "Count"
)

# Boxplot
ggplot2::ggplot(
    data = data,
    ggplot2::aes(x = Height_cm)
) +
ggplot2::geom_boxplot() +
ggplot2::labs(
    title = "Distribution of Heights",
    x = "Height (cm)"
)
```

### Categorical

- Department of everyone who works at City and Guilds
- Survey responses: Strongly disagree, disagree, Nuetral, Agree, Strongly Agree,
- Clothing sizes: Small, Medium, Large, Extra Large
- Colours: Red, Blue, Green, Yellow
- Animals: Dog, Cat, Horse, Elephant

Categorical Data can be split into two sub types: 
1. Ordinal - Categories with meaningful order, like education levels or satisfaction ratings
2. Nominal - Categories with no natural order, like department or colour.

#### Ordinal 

```{r}
ordinal_data <- data |> 
    dplyr::mutate(
        Education = factor(
            Education,
            levels = c("High School","Bachelor","Master","PhD"),
            ordered = TRUE # <- important
        )
    )

data |> 
    dplyr::count(Education)

# Bar plot
ggplot2::ggplot(
    data = ordinal_data,
    ggplot2::aes(x = Education)
) +
ggplot2::geom_bar() +
ggplot2::labs(
    title = "Distribution of Education Levels",
    x = "Education Level",
    y = "Count"
)

```

#### Nominal

```{r}
data |> 
    dplyr::count(Department)

# Bar plot
ggplot2::ggplot(
    data = ordinal_data,
    ggplot2::aes(x = Department)
) +
ggplot2::geom_bar() +
ggplot2::labs(
    title = "Distribution of Departments",
    x = "Department",
    y = "Count"
)
```

#### Ordinal vs Nominal - The R interpretation

```{r}
# Let's ensure R understands the order
ordinal_edu <- data |> 
    dplyr::mutate(
        Education = factor(
            Education,
            levels = c("High School","Bachelor","Master","PhD"),
            ordered = TRUE # <- important
        )
    ) |> 
    dplyr::pull(Education)

nominal_edu <- data |> 
    dplyr::mutate(
        Education = factor(
            Education,
            levels = c("High School","Bachelor","Master","PhD")
            #ordered = TRUE # <- important
        ),
    ) |> 
    dplyr::pull(Education)

# Why do we need to specify ordered = TRUE?
# Without this, R doesn't treat the factor as ordinal. 
# You can still "sort" the factor the same way:
sort(ordinal_edu) 
sort(nominal_edu) # like sorting a character variable alphabetically but defining a new alphabet. e.g. alphabet has order but no ranking.

# But you cannot perform numeric like operations on nominal data
ordinal_edu < "Master"
nominal_edu < "Master" # Will return error
```

## Why it matters?

- The type of data determines which visualisation or statistical test to use.
- Using the wrong statistical test could result in incorrect or misleading results.
- When you're working in R, being clear with data types can unlock different functions and analyses e.g. ordered factor vs non-ordered factor.

## Bonus data types
- Binary e.g. TRUE/FALSE, Yes/No etc.
- Date/Time
- Free text

ggplot2 cheat sheet:
Shows you what kind of visualisations you can plot with ggplot2 depending on your data type.
https://posit.co/wp-content/uploads/2022/10/data-visualization-1.pdf